The importance of transient resonances in extreme-mass-ratio inspirals

### 2016-11-08 referee_response.txt

We thank the referee for their careful reading of our document, especially given its length! We have implemented almost all of their suggestions, as indicated below, and we feel that the manuscript has been improved by these changes.

I addition to the changes commented below, we have updated some of the references and added the referee to the acknowledgements.

#1
In Fig.2, is it possible for authors to add the values of semilatus
rectum p and eccentricity e at t=t_{res}? I would like to compare the
values with those in Fig.8 as a reference.

>> Figure 8 shows the orbital parameters at plunge, rather than at resonance...

#2
In Sec.V.A the distribution of inclination angle is assume to be
uniform. Do the authors know any reference to justify this assumption?

For example, in Ref.[78], Flanagan and Hinderer (2007), Ref.[83], Sago
et al, and Ref.[84], Ganz et al, of this manuscript, it is found that
gravitational radiation reaction drives the orbital plane to
anti-align with the symmetry axis. The effect would be very small, but
it might be commented on the effect (ex. in a footnote).

>> The true inclination distribution is currently uncertain, as we are not sure of the dynamics at the very centre of galaxies. We used a uniform distribution as the least constraining proposal in absence of better information. It is true that radiation reaction will lead to anti-alignment, but we are considering the initial distribution of orbits from which inspirals begin, this is the same reason why not all the orbits are circular, despite the circularising effect of radiation reaction. Adding a comment on potential alignment is a sensible recommendation, so we have added a footnote to explain this...

#3
In the last paragraph in Sec.VI, there is the sentence ends with "if
the phase on resonance and the magnitude of the self-force can be
predicted with sufficient accuracy.".

If the authors would think that interpolation formulas of the
gravitational self-force in Kerr spacetime, which might be an
extension of Ref.[75], Warburton et al (2012), could be used for this
purpose, it might be commented in the same paragraph in Sec.VI.

#4
In Appendix B it might be mentioned that \kappa_0 in Eq.(B14),
\hat\varrho_0 in Eq.(B22), \hat\kappa_0 in Eq.(B23), \hat\kappa_{1/2}
in Eq.(B24) and \hat\varrho_{1/2} in Eq.(B25), are the constants of
integration when they appear for the first time although it might be
understood if readers would follow equations.

>> This is a good suggestion, we have added clarifications after each of the equations (or pairs of equations).

#5
I might find some typos. Please check if the authors could agree with
them:

a. In Eq.(5), parenthesis might be added to the integrand.

>> Yes, but we think it is clear as it is.

b. In the first paragraph in Sec.V.B, is "across an overlap" replaced
by "across a resonance"?

>> Yes, this has been corrected.

c. In Eq.(B19b), \hat{J_{1}} may be replaced by \hat{J_{1/2}}.

>> Yes, this has been fixed.

d. In Eq.(B29), exp(-i\pi z^2/2) may be replaced by exp(i\pi z^2/2).

>> Yes, this is extremely well spotted!

e. In Eq.(B31), 1/(is\varpi_1) may be replaced by
1/(is\varpi_1\hat{\lambda}).

>> Yes, this has been corrected.

### 2016-11-08 res_char.tex
Final Stas comment
Starting on referee comments

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits~\cite{Flanagan2012} because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum $p$ as a function of eccentricity $e$ for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentricity~\cite{ColeThesis2015}. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.

@@

We can study the effect of resonances, by comparing the adiabatic evolution to the full instantaneous evolution; a loss in SNR will reduce the number of detected events. For each inspiral, we denote the longest period of time $t\sub{ad}$ in which neither the $1$:$2$ nor the $2$:$3$ resonance is encountered. From the results of \secref{effres-phase}, we expect the recovered overlap to be approximately given by the proportion of time spent in a resonance-free region, that is $t\sub{ad} / t\sub{insp}$. This assumes that there is perfect overlap in the absence of a resonance and zero overlap across an overlap, with all times during the inspiral contributing equally. 
%In \figref{pop-adSNR-vs-Tad}, we plot the maximum overlap between the instantaneous and adiabatic models against this ratio, highlighting the number of resonances, $N\sub{res}$, each system encounters. In the majority of cases, there is no significant impact on the recovered overlap using adiabatic models.
%We can better understand the overlap distribution from \figref{pop-adSNR-vs-Tad} by considering the behaviour with the mass ratio.
In \figref{pop-adSNR-vs-eta} we plot the difference between the computed maximum overlap and the value expected from the time between resonances, highlighting the number of resonances $N\sub{res}$ each system encounters.

@@

\begin{acknowledgments}
The authors extend our sincere gratitude to \'{E}anna Flanagan and Tanja Hinderer for providing their PN self-force code, without which this work would not be possible. We are grateful to Stanislav Babak for examining both \cite{BerryThesis2013} and \cite{ColeThesis2015}, and for providing comments on this draft. We thank Tanja Hinderer, Jeandrew Brink, Maarten van de Meent, Leor Barack, Scott Hughes and Nico Yunes for useful conversations, and Christopher Moore with help preparing this manuscript. RHC was supported by STFC; CPLB thanks STFC and the Cambridge Philosophical Society; PC's work was supported by a Marie Curie Intra-European Fellowship within the 7th European Community Framework Programme (PIEF-GA-2011-299190), and JRG was supported by the Royal Society. This is LIGO Document P1600251. 
\end{acknowledgments}

@@

Using the two timescale decomposition to replace the time derivatives in the equations of motion, and substituting in the ansatz expansions gives, to first order,
\begin{widetext}
\begin{subequations}
\begin{align}
\label{eq:q-1}
\omega_0 + \eta\omega_1 + \eta\partialdiff{q_0}{\widetilde{\lambda}} + \eta\omega_0\partialdiff{q_1}{\Psi} = {} & \omega(J_0) + \eta\diff{\omega}{J}J_1 + \eta \sum_s g_s^{(1)}(J_0)\exp\left[is(\Psi + q_0)\right], \\
\eta\partialdiff{J_0}{\widetilde{\lambda}} + \eta\omega_0\partialdiff{J_1}{\Psi} = {} & \eta \sum_s G_s^{(1)}(J_0)\exp\left[is(\Psi + q_0)\right].
\label{eq:J-1}
\end{align}
\end{subequations}
\end{widetext}
Averaging \eqnref{J-1} over $\Psi$ gives\footnote{The ansatz is constructed such that $J_0 \equiv \langle J_0\rangle_\Psi$ and $q_0 \equiv \langle q_0\rangle_\Psi$.}
\begin{equation}
\partialdiff{J_0}{\widetilde{\lambda}} = G_0^{(1)}(J_0).
\label{eq:J-ad}
\end{equation}
This describes the adiabatic evolution, hence we can identify $J_0\left(\widetilde{\lambda}\right)$ with (the lowest-order piece of) the adiabatic solution~\cite{Hinderer2008}. If we similarly average \eqnref{q-1}, we find
\begin{equation}
\omega_0 + \eta\omega_1 +\eta\partialdiff{q_0}{\widetilde{\lambda}} = \omega(J_0) + \eta\partialdiff{\omega}{J}\langle J_1\rangle_\Psi + \eta g_0^{(1)}(J_0).
\end{equation}
From this we can identify the terms that originate from the frequency and, matching by order in $\eta$, obtain
\begin{subequations}
\begin{align}
\omega_0 = {} & \omega(J_0), \\
\omega_1 = {} & \partialdiff{\omega}{J}\langle J_1\rangle_\Psi.
\end{align}
\end{subequations}
This leaves
\begin{align}
\partialdiff{q_0}{\widetilde{\lambda}} = {} & g_0^{(1)}(J_0), \\
q_0 = {} & \kappa_0 + \intd{0}{\tilde{\lambda}}{g_0^{(1)}[J_0(\tau)]}{\tau}.
\label{eq:q-0-sol}
\end{align}
We now have expressions for the lowest-order terms in the expansions.

@@

Near to resonance, we consider an interior layer expansion~\cite{Kevorkian1971}. As explained in \secref{res-time}, evolution near resonance proceeds on a timescale intermediate between the slow and fast times. We therefore introduce a rescaled time
\begin{equation}
\widehat{\lambda} = \dfrac{\widetilde{\lambda} - \widetilde{\lambda}_\star}{\eta^{1/2}} = \eta^{1/2}(\lambda - \lambda_\star).
\end{equation}
As for the before resonance solution, we can create a series expansion; however, now we expand in terms of $\eta^{1/2}$~\cite{Flanagan2012}
\begin{subequations}
\begin{align}
q\left(\widehat{\lambda};\,\eta\right) = {} & \widehat{q}_0\left(\widehat{\lambda}\right) + \eta^{1/2} \widehat{q}_{1/2}\left(\widehat{\lambda}\right) + \order{\eta}, \\
J\left(\widehat{\lambda};\,\eta\right) = {} & \widehat{J}_0 + \eta^{1/2} \widehat{J}_1\left(\widehat{\lambda}\right) + \order{\eta}.
\end{align}
\end{subequations}
The series expansion for the frequency, \eqnref{omega-series}, can be rewritten as
\begin{equation}
\omega\left(\widehat{\lambda}\right) = \eta^{1/2}\varpi_1\widehat{\lambda} + \eta\varpi_2\widehat{\lambda}^2 + \order{\eta^{3/2}}.
\label{eq:omega-hat}
\end{equation}
Proceeding to write the equations of motion in terms of the rescaled time gives
\begin{subequations}
\begin{align}
\diff{q}{\widehat{\lambda}} = {} & \varpi_1\widehat{\lambda} + \eta^{1/2}\varpi_2\widehat{\lambda}^2 \nonumber \\ 
 {} & + \left. \eta^{1/2}\sum_s g_s^{(1)}\left(\widehat{J}_0,\widetilde{\lambda}_\star\right)\exp(is \widehat{q}_0)  + \order{\eta}, \right. \\
\diff{J}{\widehat{\lambda}} = {} & \eta^{1/2}\sum_s G_s^{(1)}\left(\widehat{J}_0,\widetilde{\lambda}_\star\right)\exp(is \widehat{q}_0) + \order{\eta}.
\end{align}
\end{subequations}

From the equations of motion we can match terms by their order in $\eta^{1/2}$. At zeroth order we find
\begin{equation}
\widehat{J}_0 = \widehat{\varrho}_0
\end{equation}
is constant, and
\begin{equation}
\widehat{q}_0 = \widehat{\kappa}_0 + \dfrac{\varpi_1\widehat{\lambda}^2}{2}.
\end{equation}
Using these, we can build the next-order terms
\begin{align}
\widehat{q}_{1/2} = {} & \widehat{\kappa}_{1/2} + \dfrac{\varpi_2\widehat{\lambda}^3}{3} + g_0^{(1)}(\widehat{\varrho}_0)\widehat{\lambda} \nonumber \\ 
 {} & + \sum_{s\,\neq\,0}g_s^{(1)}(\widehat{\varrho}_0)\exp(is \widehat{\kappa}_0)\intd{0}{\hat{\lambda}}{\exp\left(\dfrac{is \varpi_1\tau^2}{2}\right)}{\tau}, \\
\widehat{J}_{1/2} = {} & \widehat{\varrho}_{1/2} + G_0^{(1)}(\widehat{\varrho}_0)\widehat{\lambda} \nonumber \\
 {} & + \sum_{s\,\neq\,0}G_s^{(1)}(\widehat{\varrho}_0)\exp(is \widehat{\kappa}_0)\intd{0}{\hat{\lambda}}{\exp\left(\dfrac{is \varpi_1\tau^2}{2}\right)}{\tau}.
\end{align}
Both of these involve the complex Fresnel integral~\cite{Olver2010}, the details of which are given in the following section. We have now constructed the interior solution. % chapter 7

@@

We are interested in the asymptotic behaviour for $|\widehat{\lambda}| \rightarrow \infty$. The complex Fresnel integral has the limit~\cite{Olver2010} % [7.5.3, 7.5.4, 7.12.2, 7.12.3]
\begin{equation}
\lim_{|z|\,\rightarrow\,\infty} \mathcal{Y}(z) \sim \dfrac{\sgn z}{\sqrt{2}} \exp\left(\dfrac{i\pi}{4}\right) - \dfrac{i}{\pi z}\exp\left(-\dfrac{i\pi z^2}{2}\right),
\end{equation}
where 
\begin{equation}
\sgn z = \begin{cases}
1 & z > 0 \\
-1 & z < 0
\end{cases}\,.
\end{equation}
Hence,
\begin{align}
\lim_{|\widehat{\lambda}|\,\rightarrow\,\infty}W\left(\widehat{\lambda}\right) \sim {} & \dfrac{\sgn \widehat{\lambda}}{\sqrt{2}}\sqrt{\dfrac{\pi}{|s\varpi_1|}}\exp\left[\sgn(s\varpi_1)\dfrac{i\pi}{4}\right] \nonumber \\
  {} & + \recip{is\varpi_1}\exp\left(\dfrac{is \varpi_1 \widehat{\lambda}^2}{2}\right).
\label{eq:Fres-limit}
\end{align}

>>>>

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits~\cite{Flanagan2012} because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum $p$ as a function of eccentricity $e$ for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentricity~\cite{ColeThesis2015}. Across the grid of $a_\ast$ and $\cos\iota$ values there may be an order of magnitude variation, but the same general trends are observed. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.

@@

We can study the effect of resonances, by comparing the adiabatic evolution to the full instantaneous evolution; a loss in SNR will reduce the number of detected events. For each inspiral, we denote the longest period of time $t\sub{ad}$ in which neither the $1$:$2$ nor the $2$:$3$ resonance is encountered. From the results of \secref{effres-phase}, we expect the recovered overlap to be approximately given by the proportion of time spent in a resonance-free region, that is $t\sub{ad} / t\sub{insp}$. This assumes that there is perfect overlap in the absence of a resonance and zero overlap across a resonance, with all times during the inspiral contributing equally. 
%In \figref{pop-adSNR-vs-Tad}, we plot the maximum overlap between the instantaneous and adiabatic models against this ratio, highlighting the number of resonances, $N\sub{res}$, each system encounters. In the majority of cases, there is no significant impact on the recovered overlap using adiabatic models.
%We can better understand the overlap distribution from \figref{pop-adSNR-vs-Tad} by considering the behaviour with the mass ratio.
In \figref{pop-adSNR-vs-eta} we plot the difference between the computed maximum overlap and the value expected from the time between resonances, highlighting the number of resonances $N\sub{res}$ each system encounters.

@@

\begin{acknowledgments}
The authors extend our sincere gratitude to \'{E}anna Flanagan and Tanja Hinderer for providing their PN self-force code, without which this work would not be possible. We are grateful to Stanislav Babak for examining both \cite{BerryThesis2013} and \cite{ColeThesis2015}, and for providing comments on this draft. We thank Tanja Hinderer, Jeandrew Brink, Maarten van de Meent, Leor Barack, Scott Hughes and Nico Yunes for useful conversations, and Christopher Moore with help preparing this manuscript. We are grateful to the anonymous referee for their exceptional careful reading of the paper. RHC was supported by STFC; CPLB thanks STFC and the Cambridge Philosophical Society; PC's work was supported by a Marie Curie Intra-European Fellowship within the 7th European Community Framework Programme (PIEF-GA-2011-299190), and JRG was supported by the Royal Society. This is LIGO Document P1600251. 
\end{acknowledgments}

@@

Using the two timescale decomposition to replace the time derivatives in the equations of motion, and substituting in the ansatz expansions gives, to first order,
\begin{widetext}
\begin{subequations}
\begin{align}
\label{eq:q-1}
\omega_0 + \eta\omega_1 + \eta\partialdiff{q_0}{\widetilde{\lambda}} + \eta\omega_0\partialdiff{q_1}{\Psi} = {} & \omega(J_0) + \eta\diff{\omega}{J}J_1 + \eta \sum_s g_s^{(1)}(J_0)\exp\left[is(\Psi + q_0)\right], \\
\eta\partialdiff{J_0}{\widetilde{\lambda}} + \eta\omega_0\partialdiff{J_1}{\Psi} = {} & \eta \sum_s G_s^{(1)}(J_0)\exp\left[is(\Psi + q_0)\right].
\label{eq:J-1}
\end{align}
\end{subequations}
\end{widetext}
Averaging \eqnref{J-1} over $\Psi$ gives\footnote{The ansatz is constructed such that $J_0 \equiv \langle J_0\rangle_\Psi$ and $q_0 \equiv \langle q_0\rangle_\Psi$.}
\begin{equation}
\partialdiff{J_0}{\widetilde{\lambda}} = G_0^{(1)}(J_0).
\label{eq:J-ad}
\end{equation}
This describes the adiabatic evolution, hence we can identify $J_0\left(\widetilde{\lambda}\right)$ with (the lowest-order piece of) the adiabatic solution~\cite{Hinderer2008}. If we similarly average \eqnref{q-1}, we find
\begin{equation}
\omega_0 + \eta\omega_1 +\eta\partialdiff{q_0}{\widetilde{\lambda}} = \omega(J_0) + \eta\partialdiff{\omega}{J}\langle J_1\rangle_\Psi + \eta g_0^{(1)}(J_0).
\end{equation}
From this we can identify the terms that originate from the frequency and, matching by order in $\eta$, obtain
\begin{subequations}
\begin{align}
\omega_0 = {} & \omega(J_0), \\
\omega_1 = {} & \partialdiff{\omega}{J}\langle J_1\rangle_\Psi.
\end{align}
\end{subequations}
This leaves
\begin{align}
\partialdiff{q_0}{\widetilde{\lambda}} = {} & g_0^{(1)}(J_0), \\
q_0 = {} & \kappa_0 + \intd{0}{\tilde{\lambda}}{g_0^{(1)}[J_0(\tau)]}{\tau},
\label{eq:q-0-sol}
\end{align}
where $\kappa_0$ is the constant of integration. We now have expressions for the lowest-order terms in the expansions.

@@

Near to resonance, we consider an interior layer expansion~\cite{Kevorkian1971}. As explained in \secref{res-time}, evolution near resonance proceeds on a timescale intermediate between the slow and fast times. We therefore introduce a rescaled time
\begin{equation}
\widehat{\lambda} = \dfrac{\widetilde{\lambda} - \widetilde{\lambda}_\star}{\eta^{1/2}} = \eta^{1/2}(\lambda - \lambda_\star).
\end{equation}
As for the before resonance solution, we can create a series expansion; however, now we expand in terms of $\eta^{1/2}$~\cite{Flanagan2012}
\begin{subequations}
\begin{align}
q\left(\widehat{\lambda};\,\eta\right) = {} & \widehat{q}_0\left(\widehat{\lambda}\right) + \eta^{1/2} \widehat{q}_{1/2}\left(\widehat{\lambda}\right) + \order{\eta}, \\
J\left(\widehat{\lambda};\,\eta\right) = {} & \widehat{J}_0 + \eta^{1/2} \widehat{J}_{1/2}\left(\widehat{\lambda}\right) + \order{\eta}.
\end{align}
\end{subequations}
The series expansion for the frequency, \eqnref{omega-series}, can be rewritten as
\begin{equation}
\omega\left(\widehat{\lambda}\right) = \eta^{1/2}\varpi_1\widehat{\lambda} + \eta\varpi_2\widehat{\lambda}^2 + \order{\eta^{3/2}}.
\label{eq:omega-hat}
\end{equation}
Proceeding to write the equations of motion in terms of the rescaled time gives
\begin{subequations}
\begin{align}
\diff{q}{\widehat{\lambda}} = {} & \varpi_1\widehat{\lambda} + \eta^{1/2}\varpi_2\widehat{\lambda}^2 \nonumber \\ 
 {} & + \left. \eta^{1/2}\sum_s g_s^{(1)}\left(\widehat{J}_0,\widetilde{\lambda}_\star\right)\exp(is \widehat{q}_0)  + \order{\eta}, \right. \\
\diff{J}{\widehat{\lambda}} = {} & \eta^{1/2}\sum_s G_s^{(1)}\left(\widehat{J}_0,\widetilde{\lambda}_\star\right)\exp(is \widehat{q}_0) + \order{\eta}.
\end{align}
\end{subequations}

From the equations of motion we can match terms by their order in $\eta^{1/2}$. At zeroth order we find
\begin{equation}
\widehat{J}_0 = \widehat{\varrho}_0
\end{equation}
is constant, and
\begin{equation}
\widehat{q}_0 = \widehat{\kappa}_0 + \dfrac{\varpi_1\widehat{\lambda}^2}{2},
\end{equation}
where $\widehat{\varrho}_0$ and $\widehat{\kappa}_0$ are the constants of integration. Using these, we can build the next-order terms
\begin{align}
\widehat{q}_{1/2} = {} & \widehat{\kappa}_{1/2} + \dfrac{\varpi_2\widehat{\lambda}^3}{3} + g_0^{(1)}(\widehat{\varrho}_0)\widehat{\lambda} \nonumber \\ 
 {} & + \sum_{s\,\neq\,0}g_s^{(1)}(\widehat{\varrho}_0)\exp(is \widehat{\kappa}_0)\intd{0}{\hat{\lambda}}{\exp\left(\dfrac{is \varpi_1\tau^2}{2}\right)}{\tau}, \\
\widehat{J}_{1/2} = {} & \widehat{\varrho}_{1/2} + G_0^{(1)}(\widehat{\varrho}_0)\widehat{\lambda} \nonumber \\
 {} & + \sum_{s\,\neq\,0}G_s^{(1)}(\widehat{\varrho}_0)\exp(is \widehat{\kappa}_0)\intd{0}{\hat{\lambda}}{\exp\left(\dfrac{is \varpi_1\tau^2}{2}\right)}{\tau},
\end{align}
introducing integration constants $\widehat{q}_{1/2}$ and $\widehat{\varrho}_{1/2}$. Both the above expressions involve the complex Fresnel integral~\cite{Olver2010}, the details of which are given in the following section. We have now constructed the interior solution. % chapter 7

@@

We are interested in the asymptotic behaviour for $|\widehat{\lambda}| \rightarrow \infty$. The complex Fresnel integral has the limit~\cite{Olver2010} % [7.5.3, 7.5.4, 7.12.2, 7.12.3]
\begin{equation}
\lim_{|z|\,\rightarrow\,\infty} \mathcal{Y}(z) \sim \dfrac{\sgn z}{\sqrt{2}} \exp\left(\dfrac{i\pi}{4}\right) - \dfrac{i}{\pi z}\exp\left(\dfrac{i\pi z^2}{2}\right),
\end{equation}
where 
\begin{equation}
\sgn z = \begin{cases}
1 & z > 0 \\
-1 & z < 0
\end{cases}\,.
\end{equation}
Hence,
\begin{align}
\lim_{|\widehat{\lambda}|\,\rightarrow\,\infty}W\left(\widehat{\lambda}\right) \sim {} & \dfrac{\sgn \widehat{\lambda}}{\sqrt{2}}\sqrt{\dfrac{\pi}{|s\varpi_1|}}\exp\left[\sgn(s\varpi_1)\dfrac{i\pi}{4}\right] \nonumber \\
  {} & + \recip{is\varpi_1 \widehat{\lambda}}\exp\left(\dfrac{is \varpi_1 \widehat{\lambda}^2}{2}\right).
\label{eq:Fres-limit}
\end{align}
